---
title: "metScanR Intro"
author: "Cody Flagg"
date: "January 16, 2017"
output: html_document
---

# metScanR


## Summary 

metScanR is an R package that enables users to quickly locate and work with freely available meteorological (MET) data across multiple networks. This package can currently find data across **107,000** stations among **18** different networks across the globe. The wide range of networks and their associated, but varying documentation, meta-data, data formats, and even station identifiers can pose a major roadblock to finding, wrangling, and synthesizing MET data. 

metScanR currently allows for a user to 'bypass' many steps involved in finding MET data. A user can: 

* specify a location 
* search distance range
* sampling date ranges
* MET network (COOP, USCRN, USRCRN, ASOS, AWOS, SNOTEL, SCAN, NEON)
* meteorological variables 

metScanR will return an R list object containing all weather stations that meet the criteria. metScanR also empowers users to explore data by providing an interactive map of all returned MET stations (powered by Leaflet). 

### Version Updates: From 0.0.1 to 1.0.0 

The release of version 1.0.0 has brought not only more MET station data (from 13,000 stations to 107,000) but enhanced functionality. The expansion in MET station data required a total re-working of how data are stored and accessed. The current data set occupies a 600 megabyte binary file; this large file is loaded from a remote GitHub repo into R's 'background' when the package is loaded. 

A 10x increase in data also necessitated a new data structure, nested lists, and functions to  access those data. A series of `get...` functions are now used to quickly pull data from the metScanR_DB object. Storing data as a list rather than a data frame has two key benefits: (1) lists take up less disk space (and random access memory) and (2) lists are processed much more quickly. 

## Core Functions

* `siteFinder` : an all in one "wrapper" that accesses the functionality of several lower level `get` functions (listed below)
    * _Note_: using the `get` functions directly will return data more quickly
* `mapSiteFinder` : will map the outputs returned from `siteFinder` and/or `get` functions 
* `getCountry` : query MET stations by country of origin
* `getDates` : query by startdate, enddate, or date range (startdate AND enddate)
* `getElevation` : query by station elevation
* `getId` : find specific stations by unique identifier
* `getNearby` : find stations near a specific location and radius
* `getNetwork` : return list of stations from a given MET network
* `getVars` : query stations by environmental variables measured
    * _Note_: this function employs a 'fuzzy search' to parse through the hundreds of variables measured across MET stations. As an example, a user can narrow the granularity of their search by increasing the number of key words entered e.g. one can search for "wind", "wind speed", or "5-second wind speed". 

## Meta-Data Structure

An object, metScanR_DB, is imported into the metScanR environment when the library is loaded in R. This metScanR_DB object contains all meta-data for the MET stations captured by this project. All `metScanR` functions return meta-data in a _nested and named list_ that follows the same general structure for every station in the database. 

If we return an object named "data" from a `getVars` function call (e.g. `data <- getVars[1]`, this gives you the first list element or station of the metScanR_DB), the data are formatted as such (with a description in parentheses): 

* data$<station_id> (unique identifier) [example: data$USW00094893]
* data$<station_id>$namez 
* data$<station_id>$identifiers (associated MET networks) [example: data$USW00094893$identifiers]
  * data$<station_id>$identifiers$idType [example: data$USW00094893$identifiers$idType]
  * data$<station_id>$identifiers$id
* data$<station_id>$platform (primary MET network)
* data$<station_id>$elements (environmental variables)
  * data$<station_id>$elements$element
  * data$<station_id>$elements$date.begin
  * data$<station_id>$elements$date.end
* data$<station_id>$location
  * data$<station_id>$location$latitude_dec
  * data$<station_id>$location$longitude_dec
  * data$<station_id>$location$elev
  * data$<station_id>$location$country
  * data$<station_id>$location$state  
  * data$<station_id>$location$county
  * data$<station_id>$location$utcoffset  
  * data$<station_id>$location$date.begin
  * data$<station_id>$location$date.end  
  
TODO snapshot of one station's actual output

## Future Directions

In the near future, metScanR will provide functionality for  directly downloading MET data via existing APIs. We are also planning on including meta-data from Ameriflux and NADP stations. 

## Installation

* Install official releases from CRAN with 


```{r eval=FALSE}
install.packages("metScanR")
```


If you encounter a bug, please provide a reproducible example on this package's [github issues](https://github.com/cflagg/metScanR/issues) page. 

# Getting Started

This brief tutorial is intended for users that are both familiar and unfamiliar with R. The R code higlighted below can be copy-pasted and executed inside an R script. 

A general workflow for locating meteorological data is outlined below.

There are two primary functions that a user can interact with:  

* `siteFinder`: searches for MET stations, returns a nested list
* `mapSiteFinder`: takes a nested list returned by _siteFinder_, returns an interactive Leaflet map of all stations in the list

In this example we'll do the following:

* Locate meteorological data of interest
* Navigate metScanR's nested list structure
    * how to pull common values out of all list elements
    * how to count number of SCAN stations (`r length(getNetwork("SCAN"))`)
* Explore environmental variables listed in metScanR_DB
* Cool things you can do in metScanR v1.0.0
    * Find Rarely Monitored Variables: `getVars()` "conductivity", turbidity, ground water, gauge height
    * Map networks or variables: `getNetwork`, `mapsiteFinder(getVars("conductivity"))`
    * Map global network (SCAN)
    * Find stations with potentially correlated variables for modeling: "groundwater", "snow depth", "soil moisture" > intersect(names(list1), names(list2)) > feed into `getStation()`
    * Find the longest time series: `getDates(start = "1800", end = "2015")`
* Interactively visualize data with Leaflet
* Export the meta-data to a .CSV file

## 

## Find MET Data

### Scenario 1: locate data via latitude and longitude

Find station meta-data near a given coordinate, assign the output to object `scenario1`:

* __NOTE__: loading the metScanR library will take longer than normal (20-30 seconds) as it loads the large MET station data set into R

```{r}
library(metScanR)
scenario1 <- siteFinder(lat=40.05,lon=-105.27,startDate="2000-01-05",radius=45) # returns 40 stations
```

### Scenario 2: locate data via location, MET network, and environmental variable of interest

Our search can be narrowed to more specific variables of interest via the `vars` parameter. There are hundreds of weather/climatological variables included in the metScanR database. Rather than requiring the user to know the exact syntax, case, phrase, or network of variables before they search the database, we have implemented a 'fuzzy search' function that interprets a user's text input for the `vars` variable. The 'fuzzy search' function attempts to match the user's input against a large set of overlapping keywords for each variable listed in the database. 

In this example we search for any variables associated with air temperature measurements:

```{r}
scenario2 <- siteFinder(lat=40.05,lon=-105.27,startDate="2000-01-05",radius=45,network="COOP",vars="air temperature") # returns 15 stations
```

Typing the object name `r scenario2` shows the resulting output, where values such as "TMAX", "TMIN", and "AVGT" are returned (i.e. variables associated with air temperature measurements). Note that an error message is the resulting output if none of the stations within the search  have variables that match the `vars` argument. 

The next section walks you through how to browse through the many environmental variables logged by metScanR. 

## Explore environmental variables listed in metScanR_DB

There are 830 different environmental variables listed in the metScanR_DB. You can search through the list of terms associated with any station by combining two functions in RStudio, `View()` and plyr's `ldply`, to neatly display, and then search through, the data: 

```{r}
View(plyr::ldply(metScanR:::metScanR_terms))
```

In the search bar, typing in a term like "temperature" will narrow the number of rows displayed. 

## Navigate metScanR's nested list meta-data structure

Most R users, and data users in general, are familiar with tabular or table formatted data where information is organized in rows and columns. Because of the large volume of data contained within the metScanR database, it is necessary for the data to be stored and transferred as an R list object. The following example describes a method for manipulating metScanR's list output to quickly organize and summarize the information returned. 

Here we define a simple function that quickly extracts data values nested one "element" or two "sub-elements" within the metScanR output object. 

```{r}
## quickly extract data from the metScanR DB outputs
ls_getter <- function(input, first, second=NULL){
  if (is.null(second)){
    lapply(input, "[[", first)
  } else {
    lapply(lapply(input, "[[", first), "[[", second)
  }
}
```

With this function, and the data returned from `scenario1` above, we can begin to answer the question "What environmental variables are measured at the stations within my search radius?":

```{r}
# extract data from all stations returned
ls_getter(scenario1, "elements", "element")
```

Since every list element within the metScanR_DB has the same dimensions and data type, we can quickly coerce this 'slice' of the list (where the 'slice' is the list member and sub-member `elements$element`) into a vector by using base R functions `c`, `unique`, and `do.call`:

```{r}
## base R alternative -- outputs repeat
unique(do.call(c, ls_getter(scenario1, "elements", "element")))
```

We can also coerce the list output into a data frame using `rbind` and `do.call`: 

```{r}
do.call(rbind, ls_getter(scenario1, "identifiers", "idType"))
```

Because the number of networks is uneven across this set of stations, `do.call` "fills in" some of the blanks by repeating the vector per station. A cleaner way to assemble this data frame (that introduces an 'NA' if the vector is uneven) is to use `ldply` from the `plyr` package: 

```{r}
plyr::ldply(ls_getter(scenario1, "identifiers", "idType"), rbind)
```


## Interesting Use Cases with metScanR

### Find the oldest continuously monitoring weather station

```{r}
oldest <- getDates(startDate = "1800-01-01", endDate="2015-01-01") # search for stations that begin at the turn of the 19th century and monitor to 2015
length(oldest) # count to see how many stations metScanR finds
names(oldest) # see what the stationIDs are
oldest$USC00226177
```

### Map an entire MET network

```{r}
## Save huge map as a PDF or png to share
maps <- mapSiteFinder(getNetwork("SCAN"))
maps
```


## Interactively visualize data with Leaflet

The raw station meta-data can be viewed interactively by calling the `mapSiteFinder` function on a siteFinder object in RStudio e.g.: 

```{r}
mapSiteFinder(scenario1)
```

Users can pan, zoom, and click on stations to better visualize the spatial arrangement of stations and/or networks.  


## Write station meta-data to an external file

Since metScanR only returns station meta-data at the moment (functionality for downloading station data directly will be implemented in the future), users may want a 'hard copy' .csv of station identifiers to plug into various APIs or websites supported by MET networks: 

```{r eval=FALSE}
#write.csv(x = scenario1$finalResults, file = "path/to/your/folder/metScanR_output.csv", na="")
```

You must specify a local file path for the output via the `file` argument. 
